---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# clinspacy

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
<!-- badges: end -->

The goal of clinspacy is to perform biomedical named entity recognition, Unified Medical Language System (UMLS) concept mapping, and negation detection using the Python spaCy, scispacy, and negspacy packages.

## Installation

You can install the GitHub version of clinspacy with:

```{r eval=FALSE}
remotes::install_github('ML4LHS/clinspacy', INSTALL_opts = '--no-multiarch')
```

## Examples

```{r}
library(clinspacy)

clinspacy('This patient has diabetes and CKD stage 3 but no HTN.')

clinspacy('This patient is taking omeprazole, Protonix, and lisinopril 10 mg. He has diabetes.',
          semantic_types = 'Disease or Syndrome')

clinspacy('This patient is taking omeprazole, Protonix, and lisinopril 10 mg. He has diabetes.',
          semantic_types = 'Pharmacologic Substance')
```

## Using the mtsamples dataset

```{r}
data(mtsamples)

mtsamples[1:5,]
```


## Binding UMLS Concept Unique Identifiers to a Data Frame

This function binds columns containing concept unique identifiers with which scispacy has 99% confidence of being present with values containing frequencies. Negated concepts, as identified by negspacy's NegEx implementation, are ignored and do not count towards the frequencies.

```{r}
bind_clinspacy(mtsamples[1:5, 1:2],
               text = 'description')

bind_clinspacy(mtsamples[1:5, 1:2],
               text = 'description',
               semantic_types = 'Diagnostic Procedure')
```

## Binding Concept Embeddings to a Data Frame

```{r}
bind_clinspacy_embeddings(mtsamples[1:5, 1:2],
                          text = 'description',
                          num_embeddings = 5)

bind_clinspacy_embeddings(mtsamples[1:5, 1:2],
                          text = 'description',
                          num_embeddings = 5,
                          semantic_types = 'Diagnostic Procedure')
```

# UMLS CUI definitions

```{r}
data(cui2vec_definitions)
head(cui2vec_definitions)
```
