---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# clinspacy

<!-- badges: start -->
[![Lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
<!-- badges: end -->

The goal of clinspacy is to perform biomedical named entity recognition, Unified Medical Language System (UMLS) concept mapping, and negation detection using the Python spaCy, scispacy, and medspacy packages.

## Installation

You can install the GitHub version of clinspacy with:

```{r eval=FALSE}
remotes::install_github('ML4LHS/clinspacy', INSTALL_opts = '--no-multiarch')
```

## How to load clinspacy

```{r}
library(clinspacy)
```


## Initiating clinspacy

Initiating clinspacy is optional. If you do not initiate the package using `clinspacy_init()`, it will be automatically initiated without the UMLS linker. The UMLS linker takes up ~12 GB of RAM, so if you would like to use the linker, you can initiate clinspacy with the linker. The linker can still be added on later by reinitiating with the `use_linker` argument set to `TRUE`.

```{r}
clinspacy_init() # This is optional! The default functionality is to initiatie clinspacy without the UMLS linker
```


## Named entity recognition (without the UMLS linker)

```{r}
clinspacy('This patient has diabetes and CKD stage 3 but no HTN.')

clinspacy('HISTORY: He presents with chest pain. PMH: HTN. MEDICATIONS: This patient with diabetes is taking omeprazole, aspirin, and lisinopril 10 mg but is not taking albuterol anymore as his asthma has resolved. ALLERGIES: penicillin.')
```


## Using the mtsamples dataset

```{r}
data(mtsamples)

mtsamples[1:5,]
```

## Binding named entities to a data frame (without the UMLS linker)

Negated concepts, as identified by the medspacy cycontext flag, are ignored and do not count towards the frequencies.

```{r}
bind_clinspacy(mtsamples[1:5, 1:2],
               text = 'description')
```

## Binding entity embeddings to a data frame (without the UMLS linker)

With the UMLS linker disabled, entity embeddings can be extracted from the scispacy Python package. Up to 200-dimensional embeddings can be returned.

```{r}
bind_clinspacy_embeddings(mtsamples[1:5, 1:2],
                          text = 'description',
                          num_embeddings = 5)
```

## Adding the UMLS linker

If you would like the UMLS linker to be enabled by default, then there is no need for 

```{r}
clinspacy_init(use_linker = TRUE)
```

## Named entity recognition (with the UMLS linker)

By turning on the UMLS linker, you can restrict the results by semantic type.

```{r}
clinspacy('This patient has diabetes and CKD stage 3 but no HTN.')

clinspacy('This patient with diabetes is taking omeprazole, aspirin, and lisinopril 10 mg but is not taking albuterol anymore as his asthma has resolved.',
          semantic_types = 'Pharmacologic Substance')

clinspacy('This patient with diabetes is taking omeprazole, aspirin, and lisinopril 10 mg but is not taking albuterol anymore as his asthma has resolved.',
          semantic_types = 'Disease or Syndrome')
```

## Binding UMLS concept unique identifiers to a data frame (with the UMLS linker)

This function binds columns containing concept unique identifiers with which scispacy has 99% confidence of being present with values containing frequencies. Negated concepts, as identified by the medspacy cycontext is_negated flag, are ignored and do not count towards the frequencies.

Note that by turning on the UMLS linker, you can restrict the results by semantic type.

```{r}
bind_clinspacy(mtsamples[1:5, 1:2],
               text = 'description')

bind_clinspacy(mtsamples[1:5, 1:2],
               text = 'description',
               semantic_types = 'Diagnostic Procedure')
```

## Binding concept embeddings to a data frame (with the UMLS linker)

The default embeddings are from the scispacy R package. If you want to use the cui2vec embeddings (only available with the linker enabled), you ned to set the `type` arguement to `cui2vec`. Up to 500-dimensional embeddings can be returned.

Note that by turning on the UMLS linker, you can restrict the results by semantic type (with either type of embedding).

### Scispacy embeddings (with the UMLS linker)

With the UMLS linker enabled, you can restrict by semantic type when obtaining scispacy embeddings.

Note: The mean embeddings may be slightly different than if the linker was disabled because entities may be captured twice (as entities may map to multiple concepts).


```{r}
bind_clinspacy_embeddings(mtsamples[1:5, 1:2],
                          text = 'description',
                          num_embeddings = 5)

bind_clinspacy_embeddings(mtsamples[1:5, 1:2],
                          text = 'description',
                          num_embeddings = 5,
                          semantic_types = 'Diagnostic Procedure')
```


### Cui2vec embeddings (with the UMLS linker)

These are only available with the UMLS linker enabled.

```{r}
bind_clinspacy_embeddings(mtsamples[1:5, 1:2],
                          text = 'description',
                          type = 'cui2vec',
                          num_embeddings = 5)

bind_clinspacy_embeddings(mtsamples[1:5, 1:2],
                          text = 'description',
                          type = 'cui2vec',
                          num_embeddings = 5,
                          semantic_types = 'Diagnostic Procedure')
```

# UMLS CUI definitions

```{r}
data(cui2vec_definitions)
head(cui2vec_definitions)
```
