---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# clinspacy

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
<!-- badges: end -->

The goal of clinspacy is to perform biomedical named entity recognition, Unified Medical Language System (UMLS) concept mapping, and negation detection using the Python spaCy, scispacy, and negspacy packages.

## Installation

You can install the GitHub version of clinspacy with:

```{r eval=FALSE}
remotes::install_github('ML4LHS/clinspacy', INSTALL_opts = '--no-multiarch')
```

## How to load clinspacy

```{r}
library(clinspacy)
```


## Initiating clinspacy

Initiating clinspacy is optional. If you do not initiate the package using `clinspacy_init()`, it will be automatically initiated without the UMLS linker. The UMLS linker takes up ~12 GB of RAM, so if you would like to use the linker, you can initiate clinspacy with the linker. The linker can still be added on later by reinitiating with the `use_linker` argument set to `TRUE`.

```{r}
clinspacy_init() # This is optional! The default functionality is to initiatie clinspacy without the UMLS linker
```


## Named entity recognition (without the UMLS linker)

```{r}
clinspacy('This patient has diabetes and CKD stage 3 but no HTN.')

clinspacy('This patient with diabetes is taking omeprazole, aspirin, and lisinopril 10 mg but is not taking albuterol anymore as his asthma has resolved.')

clinspacy('This patient with diabetes is taking omeprazole, aspirin, and lisinopril 10 mg but is not taking albuterol anymore as his asthma has resolved.')
```


## Using the mtsamples dataset

```{r}
data(mtsamples)

mtsamples[1:5,]
```

## Binding named entities to a data frame (without the UMLS linker)

Negated concepts, as identified by negspacy's NegEx implementation, are ignored and do not count towards the frequencies.

```{r}
bind_clinspacy(mtsamples[1:5, 1:2],
               text = 'description')
```

## Binding entity embeddings to a data frame (without the UMLS linker)

With the UMLS linker disabled, entity embeddings can be extracted from the scispacy Python package. Up to 200-dimensional embeddings can be returned.

```{r}
bind_clinspacy_embeddings(mtsamples[1:5, 1:2],
                          text = 'description',
                          type = 'scispacy',
                          num_embeddings = 5)
```

## Adding the UMLS linker

If you would like the UMLS linker to be enabled by default, then there is no need for 

```{r}
clinspacy_init(use_linker = TRUE)
```

## Named entity recognition (with the UMLS linker)

By turning on the UMLS linker, you can restrict the results by semantic type.

```{r}
clinspacy('This patient has diabetes and CKD stage 3 but no HTN.')

clinspacy('This patient with diabetes is taking omeprazole, aspirin, and lisinopril 10 mg but is not taking albuterol anymore as his asthma has resolved.',
          semantic_types = 'Pharmacologic Substance')

clinspacy('This patient with diabetes is taking omeprazole, aspirin, and lisinopril 10 mg but is not taking albuterol anymore as his asthma has resolved.',
          semantic_types = 'Disease or Syndrome')
```

## Binding UMLS concept unique identifiers to a data frame (with the UMLS linker)

This function binds columns containing concept unique identifiers with which scispacy has 99% confidence of being present with values containing frequencies. Negated concepts, as identified by negspacy's NegEx implementation, are ignored and do not count towards the frequencies.

Note that by turning on the UMLS linker, you can restrict the results by semantic type.

```{r}
bind_clinspacy(mtsamples[1:5, 1:2],
               text = 'description')

bind_clinspacy(mtsamples[1:5, 1:2],
               text = 'description',
               semantic_types = 'Diagnostic Procedure')
```

## Binding concept embeddings to a data frame (with the UMLS linker)

The default embeddings are from the cui2vec R package. Up to 500-dimensional embeddings can be returned.

Note that by turning on the UMLS linker, you can restrict the results by semantic type.

```{r}
bind_clinspacy_embeddings(mtsamples[1:5, 1:2],
                          text = 'description',
                          num_embeddings = 5)

bind_clinspacy_embeddings(mtsamples[1:5, 1:2],
                          text = 'description',
                          num_embeddings = 5,
                          semantic_types = 'Diagnostic Procedure')
```

# UMLS CUI definitions

```{r}
data(cui2vec_definitions)
head(cui2vec_definitions)
```
